The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-06 thesis/introduction.tex
thesis; julio oct 4 comments

However, there has long been speculation that both Scl and UMi may have
been influenced by the Milky Way's tidal field. Already,
\citet{innanen+papp1979} found RR Lyrae candidate Scl members
\citep[from][]{vanagt1978} out to 180' in an elongated distribution,
speculating this to be tidal disruption. Many later density profile
determinations noted Scl's elongation and apparent outer density excess
\citetext{\citealp{eskridge1988}; \citealp{IH1995}; \citealp{walcher+2003}; \citealp{westfall+2006}; \citealp[but
see also][]{coleman+dacosta+bland-hawthorn2005}}. These works often
interpreted these features as evidence of tidal effects
\citep[e.g.,][]{walcher+2003} or sometimes a dwarf galaxy halo
\citep{westfall+2006}.

@@

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artifical disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence. For
instance, the simulations we describe later in this thesis reach the
times higher resolution than Aquarius \citep{springel+2008} with 400
times fewer particles. Idealized simulations make numerous
simplifications, neglecting mergers, cosmological context, mass
assembly, and often baryonic physics
\citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
Cosmological simulations appear to predict that tidal disruption is more
common than idealized work or the Milky Way system suggests, but the
origin and role of numerics and assumptions in this discrepancy is
unclear \citep[ \citet{riley+2024}]{panithanpaisal+2021, shipp+2023l}.
We shall use idealized simulations here to assess tidal effects after
the satellite's infall into the MW halo.

@@

To illustrate the effects of tides on an NFW halo due to a massive host,
we consider a toy model. We evolve an N-body realization of an NFW
subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
NFW static host halo with \(\rmax=25\,\kpc\) and
\(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
structure and masses of dSphs and the Milky Way, respectively. We
simulate the stellar component by assigning stellar weights to each
particle based on the relative densities of the distribution functions
in energy space (see \citet{ref:sec_painting_stars}). The stars follow
initially a 2D exponential profile with a scale radius
\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We then
evolve the orbit through a pericentre of 20 kpc starting from an
apocentre of 100 kpc. See Section~\ref{sec:methods} for a more detailed
description of our numerical setup.

>>>>

However, many works speculated that Scl and UMi were influenced by the
Milky Way's tidal field. Already, \citet{innanen+papp1979} found RR
Lyrae candidate Scl members \citep[from][]{vanagt1978} out to 180' in an
elongated distribution, speculating this to be tidal disruption. Later
density profile determinations noted Scl's elongation and apparent outer
density excess
\citetext{\citealp{eskridge1988}; \citealp{IH1995}; \citealp{walcher+2003}; \citealp{westfall+2006}; \citealp[but
see also][]{coleman+dacosta+bland-hawthorn2005}}. These works often
interpreted these features as evidence of tidal effects
\citep[e.g.,][]{walcher+2003} or sometimes a dwarf galaxy halo
\citep{westfall+2006}.

@@

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artifical disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence. For
instance, the simulations we describe later in this thesis reach the
times higher resolution than Aquarius \citep{springel+2008} with 400
times fewer particles. Idealized simulations make numerous
simplifications, neglecting mergers, cosmological context, mass
assembly, and often baryonic physics
\citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
Cosmological simulations appear to predict that tidal disruption is more
common than idealized work or the Milky Way system suggests, but the
origin and role of numerics and assumptions in this discrepancy is
unclear \citep{panithanpaisal+2021, shipp+2023, riley+2024}. We shall
use idealized simulations here to assess tidal effects after the
satellite's infall into the MW halo.

@@

To illustrate the effects of tides on an NFW halo due to a massive host,
we consider a toy model. We evolve an N-body realization of an NFW
subhalo with \(\rmax=5\,\kpc\) and \(\vmax = 27\kms\) orbiting another
NFW static host halo with \(\rmax=25\,\kpc\) and
\(\vmax = 207.4\,\kms\). These choices are motivated by the inferred
structure and masses of dSphs and the Milky Way, respectively. We
simulate the stellar component by assigning stellar weights to each
particle based on the relative densities of the distribution functions
in energy space (see Section~\ref{sec:painting_stars}). The stars follow
initially a 2D exponential profile with a scale radius
\(R_s=0.25\,\kpc\), embedded within the inner dark matter halo. We then
evolve the orbit through a pericentre of 20 kpc starting from an
apocentre of 100 kpc. See Section~\ref{sec:methods} for a more detailed
description of our numerical setup.

### 2025-10-06 thesis/results.tex
thesis; julio oct 4 comments

As discussed in Chapters \ref{sec:introduction}, \ref{sec:observations},
we test whether Galactic tides are responsible for the extended density
profiles of Scl and UMi. In this Chapter, we analyze tailored N-body
simulations, using the methods described in Chapter \ref{sec:methods},
to assess the tidal impact of the Galactic potential. To anticipate our
main conclusion, we find that tides drive dark matter loss in both
systems but leave their compact stellar components largely unaffected.
The Large Magellanic Cloud (LMC) perturbs the orbit of Scl, and somewhat
UMi, yet the resulting tidal effects are weaker than in a Milky Way-only
potential. Our simulations demonstrate that recent tides are unlikely to
have altered the stellar structure of Scl or UMi.

In this Chapter, we consider Scl first, describing tidal effects from
the MW on dark matter and stellar components. Next, we consider how
accounting for the LMC amends our conclusions. We then analyze UMi in a
similar manner, considering in turn the dark matter evolution, stellar
evolution, and orbital effects of the LMC.

@@

As a representation of an extreme tidal history, we initially
investigate the \smallperi{} Scl model.

Scl experiences moderate tidal mass loss after 10 Gyrs of evolution.
Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and
formation of diffuse streams trailing and leading Sculptor's orbit.
Because tidal stripping can be described as a gradual removal of the
least bound particles, most mass loss occurs in the outer halo.

@@

The inner density cusp is tidally resilient.
Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
velocity profiles, and the evolution of the maximum circular velocity.
The maximum velocity drops from \(31\,\kms\) to \(22\,\kms\), evolving
along the tidal track from \citet{EN2021}. The final circular velocity
profile resembles the initial with an inner cusp, but has a sharper
outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
mass, as listed in Table~\ref{tbl:scl_sim_results}. However, the inner
structure is not expected to be affected, as the Jacobi radius is over
3kpc, outside of the initial and final \(\rmax\). Thus, tides may remove
significant amounts of mass, but mostly from the outer halo.

@@

\subsection{Evolution of Sculptor's
stars}\label{evolution-of-sculptors-stars}

@@

Sculptor may not have been born as an exponential galaxy, and more
extended stellar profiles would be more sensitive to tides. As an
approximation of the present-day extended density profile, we initialize
a model with Plummer stars instead. As shown in
Fig.~\ref{fig:scl_smallperi_plummer_i_f}, the Plummer model loses more
stellar mass and forms more luminous tidal tails. Observations reaching
surface densities \(\sim10\) times fainter than our data could reveal a
stream in this case. Nevertheless, the radial extent probed by our data
remains nearly unchanged by tidal evolution.

The Jacobi and break radii further support that tidal effects should not
be apparent in the observed stellar component. As calculated for this
model, (see
Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
break and Jacobi radii both fall outside of \(\sim 100\) arcminutes for
either stellar component. Indeed, the stellar component only begins to
deviate from an exponential profile around this break radius
(Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
Given that no orbits of Scl in this potential produce significantly
smaller break or Jacobi radii, it is unlikely any other orbit would
produce a significant, observable density excess.

@@

The Milky Way isn't the only galaxy in town. Recently, work has shown
that the infall of the LMC may substantially affect the Milky Way system
\citep[e.g.,][]{erkal+2019, cautun+2019, garavito-camargo+2021, vasiliev2023}.
With a mass up to one fifth of the MW \citep[e.g.,][]{penarrubia+2015},
the LMC infall affects conclusions about the MW properties and the
orbits of satellites \citep[see
e.g.,][]{patel+2020, battaglia+2022, correamagnus+vasiliev2022}. In this
section, we examine how the LMC affects the orbital history of Sculptor.

@@

The inclusion of the LMC reshapes Scl's orbital history, as shown in
Fig.~\ref{fig:scl_lmc_orbits_effect}. In the MW-only potential, Scl's
orbit is typical of a long-term MW satellite. However, Scl's passage by
the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term orbit---Scl instead
reaches an apocentre of nearly \(300\,\kpc\). Scl may even be on first
infall, depending on the MW and LMC mass. Scl's orbit is
counter-rotating to the LMC, so Scl is unlikely to be a LMC satellite.

The timing of the LMC encounter implies a break radius consistent with
the beginning of the observed density excess. We select the orbit with
the \(6\sigma\)-smallest LMC pericentre. The orbit is selected following
the procedure in Section~\ref{sec:orbital_estimation} except with
observational uncertainties doubled. Results are similar when selecting
for a small MW pericentre instead. This \texttt{LMC-flyby} orbit, is
integrated back in time \(2\,\Gyr\) ago to isolate recent tidal effects.
We modify Scl's initial halo to have \(\rmax = 2.5\,\kpc\) and
\(\vmax = 25\,\kms\), slightly reducing the initial stellar velocity
dispersion. Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected
orbit in black and Table~\ref{tbl:orbit_ics} records the initial
conditions.

@@

The combined tidal effect of the MW and LMC are even weaker for Scl than
in the MW-only case. Fig.~\ref{fig:scl_lmc_sim_images} shows the dark
matter evolution of Scl and the passage of the LMC. With only one MW
pericentre, Scl's dark matter is less disrupted than the previous
MW-only model. The subsequent LMC passage modifies Scl's orbit but has
otherwise little effect. As a result, the dark matter structure evolves
mildly and \(\sim 50\%\) of mass remains bound
(Table~\ref{tbl:scl_lmc_sims}).

Correspondingly, the stellar component is nearly unchanged.
Fig.~\ref{fig:scl_lmc_i_f} shows the projected stellar distributions and
density profiles of this model. While the break radius is within the
observed density profile, tidal effects are too weak to be detectable.
Structural properties of the stars similarly evolve little
(Table~\ref{tbl:scl_lmc_sim_stars}).

While the instantaneous tidal force from the LMC is larger than the MW,
Scl does not experience the LMC tidal field long enough to display
disturbances. Scl and the LMC have a similar orbital plane, but orbit in
opposite directions (Fig.~\ref{fig:scl_lmc_orbits_effect}). As a result,
the interaction between Scl and the LMC is brief as their relative
velocity is \(\sim 400\,\kms\) at closest approach.

@@

We find, including only the MW potential, that tides only remove dark
matter from the outskirts of Scl. The central cusp and compact stellar
distribution are resilient to tides and any tidal effects are
well-outside the reach of current observations. We then find that the
LMC strongly perturbs Scl's orbit---in this case, Scl may be on first
infall. However, with only 1 pericentre each for the LMC and MW, the
combined tides are weaker than for our initial model. In either case, we
conclude that tides are unlikely to affect Sculptor's stellar component.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
\caption[Ursa Minor tidal tracks]{Similar to
Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
substantially more mass than Sculptor.}\label{fig:umi_tidal_track}
\end{figure}

@@

UMi only retains 3\% of its total mass after 9 Gyr
(Table~\ref{tbl:umi_sim_results}). As a result, the final dark matter
component is much smaller than the initial, but still evolves along the
predicted tidal track Fig.~\ref{fig:umi_tidal_track}. Despite the more
substantial tidal evolution, the Jacobi radius is still large, lying at
\(\sim 4\,\kpc\), well beyond the final \(\rmax\).

@@

Similar to Scl, we consider a model where UMi's stars are initially a
more extended Plummer profile, resembling the present-day density
profile. The stellar evolution of the Plummer stellar component is
similar (Fig.~\ref{fig:umi_plummer_i_f}). However, because there are
more loosely-bound stars, the Plummer model looses nearly 4\% of its
initial stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}),
and tidal features may be detectible if we measure densities 2 orders of
magnitude fainter than our present data. We show the properties of a
stream in the Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a
stream is unlikely to be observable in the near future. We conclude that
tides do not strongly affect the stellar component of this model.

@@

Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
the orbit of Ursa Minor. Predominantly, the effect is to increase the
orbital period, apocentre and pericentre. However, the orbit remains in
a similar plane and with similar morphology. As UMi is on the opposite
side as the Galaxy of the LMC, this is not unexpected.

The deviation from the MW-only orbit is mostly due to the LMC-induced
reflex motion of the Milky Way. Because the MW centre, and our origin,
is accelerated towards the LMC, UMi experiences a fictitious
acceleration opposite to the acceleration experienced by the MW centre,
causing the orbit to increase in characteristic radius.

@@

\section{Modeling uncertainties}\label{modeling-uncertainties}

@@

As Section~\ref{sec:results} shows, tides marginally affected the
stellar components of Scl and UMi, even under extreme orbits. While
Section~\ref{sec:results} only presents select models, alternative
initial conditions do not affect our qualitative conclusions on tidal
effects expected for Scl and UMi in the Galactic (and LMC) potential.

We neglect baryonic physics. Since Scl and UMi have predominantly stars
older than \(\sim 9\) Gyr
\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025},
gas dynamics are unlikely to affect recent evolution. A collisionless
dark-matter only simulation should be an excellent approximation.

Cored or less concentrated dark matter halos disrupt quicker
\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo in particular is
among the least concentrated halos consistent with UMi's velocity
dispersion. Although Scl's fiducial halo is more concentrated, less
concentrated and cored halos evolve similarly (see
Section~\ref{sec:extra_results}).

Galaxies are rarely perfect isotropic spheres. Sculptor and Ursa Minor
are elliptical, and halos are expected to be radially anisotropic
\citep[e.g.,][]{navarro+2010}. We test example non-spherical and
anisotropic models in Section~\ref{sec:extra_results}, finding that
these assumptions likely do not alter our conclusions.

A system's observed velocity dispersion directly constrains the mean
density within \(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force
required to disrupt the stellar component does not strongly depend on
the inner halo structure (via the Jacobi radius). Alternative initial
conditions may influence the total mass evolution but would produce a
similar final stellar structure.

@@

The long-term orbits of satellites are uncertain. Analytic Milky Way
potentials neglect many unknown details, including triaxiality, halo
twisting, mass evolution, and substructure. As a result, calculated
orbits may diverge significantly from the true orbits of satellites
\citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the Milky Way and
dynamical friction implies that orbits are typically less bound in the
past (less affected by tides). However, orbital energy and angular
momentum are not conserved in cosmological simulations. Orbits in
analytic potentials may overestimate the pericentre and underestimate
the maximum tidal stress \citep[although typically not by enough to
change our conclusions,][]{santistevan+2023, santistevan+2024}.

Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates the uncertainties of the
long-term orbits of Scl and UMi. Scl distance could range from
4--\(300\,\kpc\) any time more than 4Gyr ago. And, if the LMC had a
previous pericentre, UMi may have been an LMC satellite
\citep[see][]{vasiliev2024}. Notably, these potentials are still
idealized in many ways.

Motivated by Fig.~\ref{fig:scl_orbit_lmc_uncert}, we examine an extreme
pericentre of \(4\,\kpc\) of Scl with the MW in Appendix
\ref{sec:extra_results}, finding it still insufficient to produce the
observed density profile. We conclude our simulated orbits represent
reasonable extremes for \emph{recent} tidal effects. Past encounters
with the LMC are revisited below as a form of ``pre-processing'' in
Section~\ref{sec:stellar_halos}.

@@

While the long-term tidal evolution is unconstrained, we conclude that
our models are reasonable representations of recent tidal effects. As a
result, recent tides are unlikely to affect the stellar distributions of
Sculptor or Ursa Minor.

>>>>

As discussed in Chapters \ref{sec:introduction} and
\ref{sec:observations}, this thesis aims to test whether Galactic tides
are responsible for the extended density profiles of Scl and UMi. In
this Chapter, we analyze tailored N-body simulations, using the methods
described in Chapter \ref{sec:methods}, to assess the tidal impact of
the Galactic potential. To anticipate our main conclusion, we find that
tides drive dark matter loss in both systems but leave their compact
stellar components largely unaffected. The Large Magellanic Cloud (LMC)
has the potential of substantially perturbing Scl's and even UMi's
orbit, yet the resulting tidal effects are still too weak to account for
the extended outer profiles. Our simulations demonstrate that recent
tides are unlikely to have altered the stellar structure of Scl or UMi.

In this Chapter, we consider Scl first, describing tidal effects from
the MW on its dark matter and stellar components. Next, we consider how
accounting for the LMC may affect our conclusions. We then analyze UMi
in a similar manner, considering in turn the dark matter evolution,
stellar evolution, and orbital effects of the LMC.

@@

As a representation of an extreme tidal history, we initially
investigate the \smallperi{} orbit, described in
Section~\ref{sec:scl_smallperi}, chosen to maximize possible effects of
Galactic tides.

Scl experiences moderate tidal mass loss after 10 Gyrs of evolution.
Fig.~\ref{fig:scl_sim_images} shows the stripping of dark matter and the
formation of diffuse streams trailing and leading Sculptor's orbit.
Because tidal stripping can be described as a gradual removal of the
least bound particles, most mass loss occurs in the outer halo. Instead,
the inner regions of the galaxy may be relatively unaffected.

@@

The inner density cusp is tidally resilient.
Fig.~\ref{fig:scl_tidal_track} shows the initial and final circular
velocity profiles, and the evolution of the maximum circular velocity.
The maximum velocity drops from \(31\,\kms\) to \(22\,\kms\), evolving
along the tidal track from \citet{EN2021}. The final circular velocity
profile resembles the initial with an inner cusp, but has a sharper
outer truncation. Quantitatively, the halo loses \(>90\%\) its initial
mass (see Table~\ref{tbl:scl_sim_results}). However, the inner structure
is not expected to be affected, as the Jacobi radius is over 3kpc,
outside of the initial and final \(\rmax\) (see
Table~\ref{tbl:scl_sim_results} and Fig.~\ref{fig:scl_tidal_track}).
Thus, tides may remove significant amounts of mass, but mostly from the
outer halo.

@@

\subsection{Evolution of Sculptor's stars}\label{sec:scl_sim_stars}

@@

This result implies that Scl's extended profile cannot be reproduced by
Galactic tides operating on an initially exponential profile. The weak
effect of tides suggests that the outer profile of Sculptor is innate,
and not the result of tidal evolution. We check this assertion by
choosing a different initial stellar profile which matches the observed
profile and assessing how it evolves on the smallperi orbit. We show in
Fig.~\ref{fig:scl_smallperi_plummer_i_f} that a Plummer profile (instead
of an exponential) provides an adequate fit to Scl's observed profile.
The Plummer model loses more stellar mass and forms more luminous tidal
tails. Observations reaching surface densities \(\sim10\) times fainter
than our data could reveal a stream in this case. Nevertheless, over the
radial extent probed by our data, the stellar profile remains nearly
unchanged by tidal evolution.

The Jacobi and break radii further support that tidal effects should not
be apparent in the observed stellar component. As calculated for this
model, (see
Tables~\ref{tbl:scl_sim_results}, \ref{tbl:scl_sim_stars_results}), the
break and Jacobi radii both fall outside of \(\sim 100\) arcminutes for
either stellar component. Indeed, the stellar component only begins to
deviate from an exponential profile around this break radius
(Figs.~\ref{fig:scl_smallperi_i_f}, \ref{fig:scl_smallperi_plummer_i_f}).
Since no orbits of Scl produce significantly smaller break or Jacobi
radii, it is unlikely any orbit would produce an observable density
excess.

@@

The Milky Way isn't the only galaxy in town. Recently, work has shown
that the infall of the LMC may substantially affect the Milky Way system
\citep[e.g.,][]{erkal+2019, cautun+2019, garavito-camargo+2021, vasiliev2023}.
With a mass up to one fifth of the MW \citep[e.g.,][]{penarrubia+2015},
the LMC infall affects conclusions about the MW properties and the
orbits of satellites \citep[see
e.g.,][]{patel+2020, battaglia+2022, correamagnus+vasiliev2022}. In this
section, we examine how the LMC may affect the orbital history of
Sculptor.

@@

The inclusion of the LMC reshapes Scl's orbital history, as shown in
Fig.~\ref{fig:scl_lmc_orbits_effect}. In the MW-only potential, Scl's
orbit is typical of a long-term MW satellite. However, Scl's closest
approach to the LMC \(~\sim0.1\,\Gyr\) ago affects the long-term
orbit---Scl is inferred to reach an apocentre of nearly \(300\,\kpc\).
Scl may even be on first infall, depending on the MW and LMC mass. Scl's
is orbiting the Milky Way on a similar plane to the LMC, but in the
opposite direction---Scl is unlikely to be a LMC satellite.

Promisingly, the timing of the LMC encounter implies a break radius
(\(\sim 25'\), from Table~\ref{tbl:scl_lmc_sim_stars}) consistent with
the beginning of Scl's observed density excess (see
Fig.~\ref{fig:classical_dwarfs_densities}, and
Section~\ref{sec:data_density_profiles}). To probe this further, we
select an orbit with the smallest LMC-Scl pericentre (20,kpc) in the
\texttt{L3M11} model, consistent with Scl's present-day positions and
velocities. The orbit is selected following the procedure in
Section~\ref{sec:orbital_estimation} (with uncertainties doubled). This
\texttt{LMC-flyby} orbit, is integrated back in time \(2\,\Gyr\) ago to
isolate recent tidal effects. We modify Scl's initial halo to have
\(\rmax = 2.5\,\kpc\) and \(\vmax = 25\,\kms\), slightly reducing the
initial stellar velocity dispersion.
Fig.~\ref{fig:scl_lmc_orbits_effect} shows this selected orbit in black
and Table~\ref{tbl:orbit_ics} records the initial conditions.

@@

Perhaps surprisingly, the combined tidal effect of the MW and LMC are
weaker for Scl than in the MW-only case.
Fig.~\ref{fig:scl_lmc_sim_images} shows the dark matter evolution of Scl
and the passage of the LMC. With only one MW pericentre, Scl's dark
matter is less disrupted than the previous MW-only model. The subsequent
LMC passage modifies Scl's orbit but has otherwise little effect. As a
result, the dark matter structure evolves mildly and \(\sim 50\%\) of
mass remains bound (Table~\ref{tbl:scl_lmc_sims}).

JFN: This is mainly because of the altered orbit of Sculptor, which is
now inferred to come from a much larger apocentre, and, thus, to have
completed fewer orbits than in the previous MW-only ``smallperi'' orbit.
This reduces the number of pericentric passages, and therefore the net
tidal effect.

Correspondingly, the stellar component is nearly unchanged by the
combined MW and LMC tides. Fig.~\ref{fig:scl_lmc_i_f} shows the
projected stellar distributions and density profiles of this model.
While the break radius is within the observed density profile, tidal
effects are too weak to be detectable. Structural properties of the
stars similarly evolve little (Table~\ref{tbl:scl_lmc_sim_stars}).

While the instantaneous tidal force from the LMC is larger than the MW,
Scl does not experience the LMC tidal field long enough to display
disturbances. Scl and the LMC have a similar orbital plane, but orbit in
opposite directions (Fig.~\ref{fig:scl_lmc_orbits_effect}). As a result,
the interaction between Scl and the LMC is brief as their relative
velocity is \(\sim 400\,\kms\) at closest approach. Furthermore, the
Jacobi radius due to the LMC still falls outside the observed density
profile (Fig.~\ref{fig:scl_lmc_i_f}) and the MW Jacobi radius is even
larger. As a result, tides in a MW and LMC potential are even weaker
overall than for the MW-only orbit.

@@

We find, including only the MW potential, that tides only remove dark
matter from the outskirts of Scl. The central cusp and compact stellar
distribution are resilient to tides and any tidal effects are
well-outside the reach of current observations. We have also found that
the LMC strongly perturbs Scl's orbit---in this case, Scl may be on
first infall. However, with only 1 pericentre each for the LMC and MW,
the combined tides are weaker than for our initial model. In either
case, we conclude that tides are unlikely to affect Sculptor's stellar
component.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/umi_tidal_track.png}}
\caption[Ursa Minor tidal tracks]{Similar to
Fig.~\ref{fig:scl_tidal_track} except for Ursa Minor. Ursa Minor looses
substantially more mass than Sculptor. \textbf{todo: thinner and
truncated tidal track, thinner stellar
component}}\label{fig:umi_tidal_track}
\end{figure}

@@

UMi only retains 3\% of its total mass after 9 Gyr
(Table~\ref{tbl:umi_sim_results}). As a result, the final dark matter
component is much smaller than the initial, but still evolves along the
predicted tidal track (Fig.~\ref{fig:umi_tidal_track}). Despite the more
substantial tidal evolution, the Jacobi radius is still large, lying at
\(\sim 4\,\kpc\), well beyond the final \(\rmax\).

@@

The break and Jacobi radii fall well outside the observed stellar
profile. Tides would have to be far stronger to affect the observed
stellar component. As a result, the minimal tidal evolution of this
model is not unexpected.

As for Scl (Section~\ref{sec:scl_sim_stars}), we also consider a model
where UMi's stars are initially a more extended Plummer profile,
resembling the present-day density profile. The stellar evolution of
this Plummer stellar component is similar
(Fig.~\ref{fig:umi_plummer_i_f}). However, because there are more
loosely-bound stars, the Plummer model looses nearly 4\% of its initial
stellar mass to tides (Table~\ref{tbl:umi_sim_stars_results}), and tidal
features may be detectable if we measure densities 2 orders of magnitude
fainter than our present data. We show the properties of a stream in the
Appendix (Fig.~\ref{fig:umi_tidal_stream}), but such a stream is
unlikely to be observable in the near future. We conclude that tides do
not strongly affect the stellar component of this model.

@@

Fig.~\ref{fig:umi_orbits_lmc} shows the effects of including an LMC on
the orbit of Ursa Minor. Predominantly, the effect is to increase the
orbital period, apocentre and pericentre. Yet, the orbit remains in a
similar plane and with similar morphology. As UMi is on the opposite
side as the Galaxy of the LMC, and has a closest LMC approach of
\(\gtrsim 100\,\kpc\), this is not surprising.

The deviation from the MW-only orbit is mostly due to the LMC-induced
reflex motion of the Milky Way. Because the MW centre is accelerated
towards the LMC and away from UMi, UMi's orbit to increase in
characteristic radius.

@@

\section{Modelling uncertainties}\label{modelling-uncertainties}

@@

As the above results show, tides only marginally affected the stellar
components of Scl and UMi, even orbit chosen to have the smallest
observationally-consistent pericentres. While we have only presented
select models, alternative initial conditions do not affect our
qualitative conclusions on tidal effects expected for Scl and UMi in the
Galactic (and LMC) potential.

Although our analysis neglects baryonic physics, Scl and UMi have
predominantly stars older than \(\sim 9\) Gyr
\citep{carrera+2002, deboer+2011, weisz+2014, delosreyes+2022, sato+2025}.
So, gas dynamics are unlikely to affect recent evolution. A
collisionless dark-matter only simulation should therefore be an
excellent approximation.

Cored or less concentrated dark matter halos disrupt quicker
\citep[e.g.,][]{stucker+2023}. Our fiducial UMi halo in particular is
among the least concentrated halos consistent with UMi's velocity
dispersion. Although Scl's fiducial halo is more concentrated, less
concentrated and cored halos evolve similarly (see Appendix
\ref{sec:extra_results}).

Galaxies are rarely perfect isotropic spheres. Sculptor and Ursa Minor
are elliptical, and halos are expected to be radially anisotropic
\citep[e.g.,][]{navarro+2010}. We test non-spherical and anisotropic
models in Appendix \ref{sec:extra_results}, finding that these
assumptions likely do not alter our conclusions.

Alternative initial conditions may influence the total mass evolution
but should produce a similar final stellar structure. A system's
observed velocity dispersion directly constrains the mean density within
\(R_h\) \citep[e.g.,][]{wolf+2010}. Thus, the tidal force required to
disrupt the stellar component does not strongly depend on the inner halo
structure (via the Jacobi radius).

@@

The long-term orbits of satellites are uncertain. Analytic Milky Way
potentials neglect many unknown details, including triaxiality, halo
twisting, mass evolution, and substructure. Due to these inadequacies,
calculated orbits may diverge significantly from the true orbits of
satellites \citep[e.g.,][]{dsouza+bell2022}. The mass-growth of the
Milky Way and dynamical friction imply that orbits are typically less
bound in the past (less affected by tides). Orbital energy and angular
momentum of subhalos are not conserved in cosmological N-body
simulations. Consequently, orbits in analytic potentials may
overestimate the pericentre and underestimate the maximum tidal stress
\citep[although typically not by enough to change our
conclusions,][]{santistevan+2023, santistevan+2024}.

As an example, Fig.~\ref{fig:scl_orbit_lmc_uncert} illustrates how
changes to the LMC potential modifies the long-term orbital trajectories
of Scl and UMi. More than 4 Gyr ago, orbits of Scl diverge
substantially. Some orbits are near apocentres of \(\sim 300\,\kpc\)
when others approach pericentres as small as \(\sim 10\,\kpc\). Ursa
Minor's orbit is more stable until the possible previous LMC pericentre.
In some cases, Ursa Minor may have been bound to the LMC.

Motivated by Fig.~\ref{fig:scl_orbit_lmc_uncert}, we examine an extreme
pericentre of \(4\,\kpc\) of Scl with the MW in Appendix
\ref{sec:extra_results}, finding it still insufficient to produce the
observed density profile. Regardless, we conclude our simulated orbits
represent reasonable extremes for \emph{recent} tidal effects. Past
encounters with the LMC are revisited below as a form of
``pre-processing'' in Section~\ref{sec:stellar_halos}.

@@

While the long-term tidal evolution is unconstrained, we conclude that
our models are reasonable representations of recent tidal effects. As a
result, recent tides are unlikely to affect the stellar distributions of
Sculptor and Ursa Minor.
